Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-12 _paper.tex
-Penguin, Chocolate, and Ames update -Introduce PI/CI, Obs->inst, var->feat

\begin{document}
\maketitle
\begin{abstract}
The increased predictive power comes at the cost of interpretability, which has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models are using predictors, to arrive at a prediction, with a point estimate of the linear variable importance in the vicinity of each observation. These can be considered to be linear projections, and as such can be further explored interactively to better understand the interaction between variables used to make predictions, across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any observation, and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (football salaries, house prices) response variables. The methods are implemented in the \textbf{R} package \textbf{cheem}, available on CRAN.
\end{abstract}

@@

There are different reasons and emphases to fit a model. Breiman (2001), reiterated by Shmueli (2010), taxonomizes modeling based on its purpose; \emph{explanatory} modeling is done for some inferential purpose, while \emph{predictive} modeling focuses more on the predictions of out-of-sample observations. The intended use has important implications for model selection and development. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for more accurate non-linear models. The use of black-box models is becoming increasingly common, but not without their share of controversy (O'Neil 2016; Kodiyan 2019). However, the loss of interpretation presents a challenge.

Interpretability is vital for exploring and protecting against potential biases (e.g.~sex (Dastin 2018; Duffy 2019), race (Larson et al. 2016), and age (Díaz et al. 2018)) in any model. For instance, models regularly pick up on biases in the training data that have observed influence on the response (output) variable, which is then built into the model. Variable-level (feature-level) interpretability of models is essential in the evaluation models for such biases. It is also generally important for many problems, where it is not enough to accurately predict accurately but one must be able to explain which predictors are most responsible in generating a response value.

Another concern is that of data drift, which is a shift in support or domain of the explanatory variables (feature or predictors). Non-linear models are typically more sensitive, and do not extrapolate well outside of the training data domain. Better interpretability of the model means that there is more transparency where models' predictions may be plausible or completely unreliable.

Explainable Artificial Intelligence (XAI) is an emerging field of research that tries to increase the interpretability of black-box models. A common approach is to use \emph{local explanations}, which attempt to approximate linear variable importance at the location of each observation (instance), or the predictions at a specific point in the data domain. Because these are point specific, a challenge is visualizing them to more comprehensively understand a model.

In multivariate data visualization, a \emph{tour} (Asimov 1985; Buja and Asimov 1986; S. Lee et al. 2021) is a sequence of linear projections of data onto a lower-dimensional space. Tours are viewed as an animation over minor changes to the projection basis. Structure in a projection can then be explored visually to see which variables contribute to the formation of that structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the shape of the shadow change conveys information of the structure and features of the object.

There are various types of tours distinguished by the generation of projection bases. In a \emph{manual} tour (Cook and Buja 1997; Spyrison and Cook 2020), the path is defined by changing the contribution of a selected variable. Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries (Wickham, Cook, and Hofmann 2015), and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis paths (Y. D. Lee et al. 2013; da Silva, Cook, and Lee 2021).

In our proposed approach we use the radial manual tour to scrutinize a local explanation. Additional interactivity allows the user to identify an observation of interest, then explore its local explanation by changing variable contribution with the radial tour. The methods are implemented in R package \textbf{cheem}. Example datasets are provided to illustrate usage for classification and regression tasks.

Using a radial tour can be considered similar to counterfactual, what-if analysis, such as \emph{ceteris paribus} (Biecek 2020). This phrase, Latin for ``other things held constant'' or ``all else unchanged'', shows how an observation's prediction would change from a marginal change in one explanatory variable given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation of the factual; it varies contributions of the variables by rotating the basis, a reorientation of the data object. A constraint in our approach is that the basis must remain orthonormal. That means, when the contribution of one variable decreases, the contributions of others necessarily increase such that there is a complete component in that direction. This also ensures that what is seen is strictly a low-dimensional projection from high-dimensions, and is thus an interpretable visualisation.

The remainder of this paper is organized as follows. The following Section, \ref{sec:explanations} covers the background of the local explanation, and the traditional visuals produced. Section \ref{sec:tour} explains the animations of continuous linear projections. Section \ref{sec:applicationdesign} discusses the visual layout in the interactive interface, how they facilitate analysis, data preprocessing, and package infrastructure. Then Section @ref(\#sec:casestudies) illustrates the application to supervised learning with categorical and quantitative response variables. We conclude with Section \ref{sec:cheemdiscussion} of the insights gained and directions that might be explore in the future.

@@

Consider a highly non-linear model. It can be hard to determine whether small changes in a variable's value will make a class prediction change group or identify which variables contribute to an extreme residual. Local explanations shed light on these situations by approximating linear variable importance in the vicinity of a single observation or point.

@@

These instance-level explanations are used in various ways depending on the data. In image classification, where pixels would correspond to predictors, saliency maps overlay or offset a heatmap indicating important pixels (Simonyan, Vedaldi, and Zisserman 2014). For instance, pixels corresponding to snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words (Vanni et al. 2018). In the case of numeric regression, they are used to explain variable additive contributions from the model intercept to the observation's prediction (Ribeiro, Singh, and Guestrin 2016).

SHaply Additive exPlanations (SHAP) approximates the variable importance in the vicinity of one observation conceptually by examining the effect of other variables on the contribution of the variable of interest on predicting the response. This explanations almost all point to Shapley (1953)'s method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score. Strumbelj and Kononenko (2010) introduced the use of SHAP for local explanations in ML models. While the models themselves are invariant to the order of the predictors, the additive explanations are not. The SHAP values are the mean contributions over some subset these permutations. This is also related to partial dependence plots (Molnar 2020), used to explain the effect of a variable by predicting the response for a range of values on this variable, after fixing the value of all other variables to their mean. Partial dependence plots are a global approximation of the variable importance, while SHAP is specific to one instance. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in Wickham, Cook, and Hofmann (2015), which helps to understand the relative importance of each variable in the context of all other candidate variables.

For our application, we use \emph{tree SHAP}, a variant of SHAP enjoys a lower computational complexity (Lundberg, Erion, and Lee 2018). Instead of aggregating over permutations of the variable order, tree SHAP approximates instance level feature importance by exploring the stucture of the decision trees. Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized explanations as the starting projection basis (call this the \emph{attribution projection}) to further scrutinize the explanation.

@@

\caption{Illustration of the use of SHAP values for a random forest model for salaries of FIFA 2020 players based on nine predictors corresponding to different skills. A star offensive and densive player, Messi and a defensive player are compared, L. Messi and V van Dijk respectively. Plot a) is the break-down plot showing one additive variable-attributution. These attributions are sensitive to the order of the variables. Plot b) shows the distribution of attribution for each variable across 25 such permutation of the order of predictors, with the mean displayed as a dot, for each players. Offense and movement are important for Messi but not van Dijk, and conversely defense and power are important for van Dijk but not Messi.}\label{fig:shapdistrbd}
\end{figure}

Following the use case \emph{Explanatory Model Analysis} (Biecek and Burzykowski 2021), we use FIFA data to illustrate the use of SHAP. Consider soccer data from the FIFA 2020 season (Leone 2020). There are 5000 observations of 9 skill measures (after aggregating highly correlated variables). A random forest model is fit regressing wages {[}2020 Euros{]}, from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). The results are displayed in Figure \ref{fig:shapdistrbd}. We expect to see a difference in the attribution of the variable importance across the two positions of the players, which would be interpreted as how the player's salary depends on this combination of skill sets. Plot (b) is a modified breakdown plot (Gosiewska and Biecek 2019) where the order of variables is fixed, so the two observations can be more easily compared.

In summary, these plots highlight how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this instance, two players with different positions receive different profiles of variable importance to explain the prediction of their wages.

@@

A \emph{tour} enables viewing of high-dimensional data by animating many linear projections with small incremental changes. It is achieved by following a path of linear projections (bases) of high-dimensional space. One of the features of the tour is the object permanence of the data points; one can track the relative change of observations in time, and as such gain information about the relationships between points across multiple variables. There are various types of tours that are distinguished by how the paths are generated (S. Lee et al. 2021; Cook et al. 2008).

The manual tour (Cook and Buja 1997) defines its path by changing a selected variable's contribution to a basis, to allow the variable to contribute more or less to the projection. The contribution of all other variables is constrained by the requirement that a basis needs to be orthonormal (column correspond to vectors, with unit length, and orthogonal to each other). The manual tour is primarily used to assess the importance of a variable to structure visible in a projection. It also lends itself to pre-computation to be queued in advance or computed on-the-fly for human-in-the-loop analysis (Karwowski 2006).

A version of the manual tour called a \emph{radial tour} is implemented in Spyrison and Cook (2020) and forms the basis of the new work. In a radial tour, the selected variable is allowed to change its magnitude of contribution but not its angle; it must move along the direction of its original contribution. The implementation allows for pre-computation and also interactive re-calculation to focus on a different variable.

@@

To explore the local explanations, an ensemble of plots (Unwin and Valero-Mora 2018) is provided, called the \emph{cheem viewer}. There are two primary plots: the global view to give the context of all of the SHAP values, and the radial tour view to explore the local explanations with user-controlled rotation. In addition, there are numerous user inputs, including variable selection for the radial tour, and instance selection for making comparisons. Figures \ref{fig:classificationcase} and \ref{fig:regressioncase} contain screenshots showing the cheem viewer for the two primary tasks classification (categorical response) and regression (quantitative response).

@@

The global view provides the context of all observations and facilitates the exploration of the separability of the data- and attribution-spaces. Both of these spaces are of dimension \(n\times p\), where \(n\) is the number of observations and \(p\) is the number of predictors. The attribution space corresponds to the local explanations for each observation, which will have \(p\) values for each observation.

A visualisation of these spaces is provided by the first two principal components of their respective spaces. In addition, a plot observed by predicted response is also provided. In both PCA plots the orientation and magnitude of the variables are inscribed on a unit circle, similar to a biplot (Gabriel 1971). A single 2D projection will not encompass all of the structure of higher-dimensional space, but it is generally a useful visual summary. For classification tasks, misclassified observations are circled in red if applicable. Linked brushing between the plots is provided and a tabular display of selected points helps to facilitate exploration of the spaces and the model.

While the comparison of these spaces is interesting, a main purpose of the global view is to enable the selection of observations, from which to explore the local explanations.

@@

The global view facilitated the selection of a primary and optional comparison observation. The variable-level attribution of the primary observation is normalized and used as the initial 1D basis in a radial tour. This is an approximation of the contributions of the linear variables that best explain the difference between the model intercept and an observation's prediction, not the local shape of the model surface.

The initial frame is the normalized SHAP values of the primary observation. The current projection basis is depicted as the width of a bar, the variable's contribution to the horizontal axis. The normalized values of all observations are shown as vertical parallel coordinate plots.

The radial tour creates a basis path by varying the contribution of a selected variable, fully into and out of a projection frame. Doing so tests an individual variable's sensitivity to the structure identified by the local explanation. The default variable selected has the largest discrepancy between the attribution of primary and comparison observations. The following sections elaborate on the takeaways from applying this approach in classification and regression tasks. Now that we have introduced the global view and corresponding cheem radial tour, let us discuss the differences between the classification and regression cases.

@@

Typically we select a misclassified observation compared to a correctly classified point nearby in data space. The model infomation in the global view is a depiction of the model conmfusion matrix. The radial tour is 1D, with density display, while the goal of the tour; exploring the sensitivity of each variable to structure identified by the local explanation, evaluating the support or robustness of the prediction.

@@

\caption{Overview of the cheem viewer for classification tasks. Plots are colored on predicted class, and red circles indicate misclassified observations. Model visual is essentially a representation of the confusion matrix of the model. The radial tour is a 1D projection starting the attribution projection. The first frame is the linear variable importances that best describe the observation's prediction. We probe the support of variable contributions by selecting a variable and use the radial tour to vary its coefficient.}\label{fig:classificationcase}
\end{figure}

@@

A tooltip displays observation number/name and classification information while the cursor hovers over a point. Linked brushing allows the selection of points (left click and drag) where those points will be highlighted across plots. The information corresponding to the selected points is populated on a dynamic table. These interactions aid exploration of the spaces and, finally, identification of a primary and comparison observation.

@@

\caption{Illustration of data explorations interactions in the global view. This view has linked brushing, where observations selected in one facet are highlighted in the other facets and populate an interactive tabular display below. Tooltips display when hovering over an observation.}\label{fig:interactions}
\end{figure}

@@

\begin{itemize}
    \item \textbf{Data:} predictors and response are unscaled complete numerical matrix. Most models and local explanations are scale invariant.
    \item \textbf{Model and explanation:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [Liaw and Wiener (2002)], compatibility tree SHAP. We use modest hyperparameters, namely: 125 trees, number variables randomly sampled at each split, mtry = $\sqrt{p}$ or $p/3$ for classification and regression, and minimum size of terminal nodes $max(1, n/500)$ or $max(5, n/500)$ for classification and regression. Tree SHAP is calculated for \emph{each} observation using the package \textbf{treeshap} Kominsarczyk et al. (2021). This implementation aggregates over exhaustively over the attribution of all trees and we opt to not to fit interactions of features. 
    \item \textbf{Cheem view:} after the model and full explanation space are calculated we scale each feature by standard deviations away from the mean to achieve a common support for visuals. Statistics for mapping to color are calculated on the scaled spaces. Interactive tabular display reports the original values.
\end{itemize}

The time to preprocess the data will vary significantly with the model and local explanation. For reference, the FIFA data, 5000 observations of nine explanatory variables, took 2.5 seconds to fit a random forest model of modest hyperparameters. Extracting the tree SHAP values of each observation took 270 seconds combined. PCA and statistics of the variables and attributions took 2.8 seconds. These runtimes were from a non-parallelized R session on a modern laptop, but suffice to say that the bulk of the time will be spent on the local attribution. An increase in model complexity or data dimensionality will quickly become an obstacle. With its reduced computational complexity, this makes tree SHAP a good candidate to start with. Alternatively, the package \textbf{fastshap} (Greenwell 2020) claims extremely low runtimes, which are attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting.

@@

The Palmer penguins data (Gorman, Williams, and Fraser 2014; Horst, Hill, and Gorman 2020) was collected on three species of penguins foraging near Palmer Station, Antarctica. The data was publicly available to be a substitute for the overly-used iris data, and is quite similar in form. The subset used here has missing values removed (as used in the \texttt{spinifex} package), and there are 333 observations and four physical measurements, \texttt{bill\_length\_mm} (\texttt{b\_l}), \texttt{bill\_depth\_mm} (\texttt{b\_d}), \texttt{flipper\_length\_mm} (\texttt{f\_l}), \texttt{body\_mass\_g} (\texttt{wgt}), for this illustration. A random forest model was fitted with species as the response variable.

@@

\caption{Examining the SHAP values for a random forest model classifying Palmer penguin species. The observation of interest is an Adelie (green) penguin that is misclassified as a Chinstrap (orange), marked as an asterisk in (a), and the dashed vertical line in (b). XXX needs fixing Starting from the attribution projection the contribution of \texttt{b\_l} is varied. This explanation is not supported sense by the time there is moderate contribution. The current basis is fairly close to the regularly attribution to other chinstrap penguins aside from the sign on flipper length being flipped. It feels odd that SHAP has identified a profile where chinstrap and adelie penguins are hard to distinguish when there are separating orientations that should be found within 125 trees. The .mp4 animation of this tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{shorturl.at/hFHNS}}\label{fig:casepenguins}
\end{figure}

Figure \ref{fig:casepenguins} shows plots from the cheem viewer for exploring the random forest model on the penguins data. Plot (a) shows the global view, and plot (b) shows several 1D projections generated with the radial tour. Penguin 124, an Adelie (green), is the observation of interest because it has been misclassified as a Chinstrap (orange).

There is more separation visible in the attribution space than the data space, as would be expected. The observed vs predicted plot reveals a handful of misclassified observations. A misclassified point (represented by the asterisk in the global view, and as a dashed vertical line in the tour view) is contrasted with a correctly classified point (represented by an \(\times\) and vertical dotted line).

The radial tour starts from the attribution projection of the misclassified observation (left plot). The important variables identifed by SHAP in the (wrong) prediction for this observation is mostly \texttt{b\_d} with smaller contributions of \texttt{f\_l} and \texttt{wgt}. This projection is a view where the Gentoo (purple) penguins are most distinguished from Adelie and Chinstrap. For this observation we would like to know why an Adelie has been wrongly labelled a Chinstrap, so this combination of variables is not particularly useful. To explore this we use the radial tour to vary the contribution of bill length (\texttt{b\_l}). The more \texttt{b\_l} contributes to the the projection the more this penguin looks like a Chinstrap, that it is mislabelled as. This suggests that \texttt{b\_l} should have been an important variable for explaining the (wrong) prediction.

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales across nine variables. Using interaction from the global view, we select a house with an extreme negative residual and an accurate observation close to it in the data.

>>>>

\begin{document}
\maketitle
\begin{abstract}
The increased predictive power comes at the cost of interpretability, which has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models are using predictors, to arrive at a prediction, with a point estimate of the linear feature importance in the vicinity of each instance. These can be considered to be linear projections, and as such can be further explored interactively to better understand the interaction between features used to make predictions, across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any instance, and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (football salaries, house prices) response features. The methods are implemented in the \textbf{R} package \textbf{cheem}, available on CRAN.
\end{abstract}

@@

There are different reasons and emphases to fit a model. Breiman (2001), reiterated by Shmueli (2010), taxonomizes modeling based on its purpose; \emph{explanatory} modeling is done for some inferential purpose, while \emph{predictive} modeling focuses more on the predictions of out-of-sample instances. The intended use has important implications for model selection and development. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for more accurate non-linear models. The use of black-box models is becoming increasingly common, but not without their share of controversy (O'Neil 2016; Kodiyan 2019). However, the loss of interpretation presents a challenge.

Interpretability is vital for exploring and protecting against potential biases (e.g.~sex (Dastin 2018; Duffy 2019), race (Larson et al. 2016), and age (Díaz et al. 2018)) in any model. For instance, models regularly pick up on biases in the training data that have observed influence on the response (output) feature, which is then built into the model. Feature-level (variable-level) interpretability of models is essential in the evaluation models for such biases. It is also generally important for many problems, where it is not enough to accurately predict accurately but one must be able to explain which predictors are most responsible in generating a response value.

Another concern is that of data drift, which is a shift in support or domain of the explanatory features (variable or predictors). Non-linear models are typically more sensitive, and do not extrapolate well outside of the training data domain. Better interpretability of the model means that there is more transparency where models' predictions may be plausible or completely unreliable.

Explainable Artificial Intelligence (XAI) is an emerging field of research that tries to increase the interpretability of black-box models. A common approach is to use \emph{local explanations}, which attempt to approximate linear feature importance at the location of each instance (observation), or the predictions at a specific point in the data domain. Because these are point specific, a challenge is visualizing them to more comprehensively understand a model.

In multivariate data visualization, a \emph{tour} (Asimov 1985; Buja and Asimov 1986; S. Lee et al. 2021) is a sequence of linear projections of data onto a lower-dimensional space. Tours are viewed as an animation over minor changes to the projection basis. Structure in a projection can then be explored visually to see which features contribute to the formation of that structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the shape of the shadow change conveys information of the structure and features of the object.

There are various types of tours distinguished by the generation of projection bases. In a \emph{manual} tour (Cook and Buja 1997; Spyrison and Cook 2020), the path is defined by changing the contribution of a selected feature. Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries (Wickham, Cook, and Hofmann 2015), and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis paths (Y. D. Lee et al. 2013; da Silva, Cook, and Lee 2021).

In our proposed approach we use the radial manual tour to scrutinize a local explanation. Additional interactivity allows the user to identify an instance of interest, then explore its local explanation by changing feature contribution with the radial tour. The methods are implemented in R package \textbf{cheem}. Example datasets are provided to illustrate usage for classification and regression tasks.

Using a radial tour can be considered similar to counterfactual, what-if analysis, such as \emph{ceteris paribus} (Biecek 2020). This phrase, Latin for ``other things held constant'' or ``all else unchanged,'' shows how an instance's prediction would change from a marginal change in one explanatory feature given that other features are held constant. It ignores correlations of the features and imagines a case that was not observed. In contrast, our approach is a geometric explanation of the factual; it varies contributions of the features by rotating the basis, a reorientation of the data object. A constraint in our approach is that the basis must remain orthonormal. That means, when the contribution of one feature decreases, the contributions of others necessarily increase such that there is a complete component in that direction. This also ensures that what is seen is strictly a low-dimensional projection from high-dimensions, and is thus an interpretable visualisation.

The remainder of this paper is organized as follows. The following Section, \ref{sec:explanations} covers the background of the local explanation, and the traditional visuals produced. Section \ref{sec:tour} explains the animations of continuous linear projections. Section \ref{sec:applicationdesign} discusses the visual layout in the interactive interface, how they facilitate analysis, data preprocessing, and package infrastructure. Then Section @ref(\#sec:casestudies) illustrates the application to supervised learning with categorical and quantitative response features. We conclude with Section \ref{sec:cheemdiscussion} of the insights gained and directions that might be explore in the future.

@@

Consider a highly non-linear model. It can be hard to determine whether small changes in a feature's value will make a class prediction change group or identify which features contribute to an extreme residual. Local explanations shed light on these situations by approximating linear feature importance in the vicinity of a single instance.

@@

These instance-level explanations are used in various ways depending on the data. In image classification, where pixels would correspond to predictors, saliency maps overlay or offset a heatmap indicating important pixels (Simonyan, Vedaldi, and Zisserman 2014). For instance, pixels corresponding to snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words (Vanni et al. 2018). In the case of numeric regression, they are used to explain feature additive contributions from the model intercept to the instance's prediction (Ribeiro, Singh, and Guestrin 2016).

SHaply Additive exPlanations (SHAP) approximates the feature importance in the vicinity of one instance conceptually by examining the effect of other features on the contribution of the feature of interest on predicting the response. This explanations almost all point to Shapley (1953)'s method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score. Strumbelj and Kononenko (2010) introduced the use of SHAP for local explanations in ML models. While the models themselves are invariant to the order of the predictors, the additive explanations are not. The SHAP values are the mean contributions over some subset these permutations. This is also related to partial dependence plots (Molnar 2020), used to explain the effect of a feature by predicting the response for a range of values on this feature, after fixing the value of all other features to their mean. Partial dependence plots are a global approximation of the feature importance, while SHAP is specific to one instance. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in Wickham, Cook, and Hofmann (2015), which helps to understand the relative importance of each feature in the context of all other candidate features.

For our application, we use \emph{tree SHAP}, a variant of SHAP enjoys a lower computational complexity (Lundberg, Erion, and Lee 2018). Instead of aggregating over permutations of the feature order, tree SHAP approximates instance level feature importance by exploring the stucture of the decision trees. Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized explanations as the starting projection basis (call this the \emph{attribution projection}) to further scrutinize the explanation.

@@

\caption{Illustration of the use of SHAP values for a random forest model for salaries of FIFA 2020 players based on nine predictors corresponding to different skills. A star offensive and densive player, Messi and a defensive player are compared, L. Messi and V van Dijk respectively. Plot a) is the break-down plot showing one additive feature-attributution. These attributions are sensitive to the order of the features. Plot b) shows the distribution of attribution for each feature across 25 such permutation of the order of predictors, with the mean displayed as a dot, for each players. Offense and movement are important for Messi but not van Dijk, and conversely defense and power are important for van Dijk but not Messi.}\label{fig:shapdistrbd}
\end{figure}

Following the use case \emph{Explanatory Model Analysis} (Biecek and Burzykowski 2021), we use FIFA data to illustrate the use of SHAP. Consider soccer data from the FIFA 2020 season (Leone 2020). There are 5000 instances of 9 skill measures (after aggregating highly correlated features). A random forest model is fit regressing wages {[}2020 Euros{]}, from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). The results are displayed in Figure \ref{fig:shapdistrbd}. We expect to see a difference in the attribution of the feature importance across the two positions of the players, which would be interpreted as how the player's salary depends on this combination of skill sets. Plot (b) is a modified breakdown plot (Gosiewska and Biecek 2019) where the order of features is fixed, so the two instances can be more easily compared.

In summary, these plots highlight how local explanations bring interpretability to a model, at least in the vicinity of their instances. In this instance, two players with different positions receive different profiles of feature importance to explain the prediction of their wages.

@@

A \emph{tour} enables viewing of high-dimensional data by animating many linear projections with small incremental changes. It is achieved by following a path of linear projections (bases) of high-dimensional space. One of the features of the tour is the object permanence of the data points; one can track the relative change of instances in time, and as such gain information about the relationships between points across multiple features. There are various types of tours that are distinguished by how the paths are generated (S. Lee et al. 2021; Cook et al. 2008).

The manual tour (Cook and Buja 1997) defines its path by changing a selected feature's contribution to a basis, to allow the feature to contribute more or less to the projection. The contribution of all other features is constrained by the requirement that a basis needs to be orthonormal (column correspond to vectors, with unit length, and orthogonal to each other). The manual tour is primarily used to assess the importance of a feature to structure visible in a projection. It also lends itself to pre-computation to be queued in advance or computed on-the-fly for human-in-the-loop analysis (Karwowski 2006).

A version of the manual tour called a \emph{radial tour} is implemented in Spyrison and Cook (2020) and forms the basis of the new work. In a radial tour, the selected feature is allowed to change its magnitude of contribution but not its angle; it must move along the direction of its original contribution. The implementation allows for pre-computation and also interactive re-calculation to focus on a different feature.

@@

To explore the local explanations, an ensemble of plots (Unwin and Valero-Mora 2018) is provided, called the \emph{cheem viewer}. There are two primary plots: the global view to give the context of all of the SHAP values, and the radial tour view to explore the local explanations with user-controlled rotation. In addition, there are numerous user inputs, including feature selection for the radial tour, and instance selection for making comparisons. Figures \ref{fig:classificationcase} and \ref{fig:regressioncase} contain screenshots showing the cheem viewer for the two primary tasks: classification (categorical response) and regression (quantitative response).

@@

The global view provides the context of all instances and facilitates the exploration of the separability of the data- and attribution-spaces. Both of these spaces are of dimension \(n\times p\), where \(n\) is the number of instances and \(p\) is the number of predictors. The attribution space corresponds to the local explanations for each instance, which will have \(p\) values for each instance.

A visualisation of these spaces is provided by the first two principal components of their respective spaces. In addition, a plot observed by predicted response is also provided. In both PCA plots the orientation and magnitude of the features are inscribed on a unit circle, similar to a biplot (Gabriel 1971). A single 2D projection will not encompass all of the structure of higher-dimensional space, but it is generally a useful visual summary. For classification tasks, misclassified instances are circled in red if applicable. Linked brushing between the plots is provided and a tabular display of selected points helps to facilitate exploration of the spaces and the model.

While the comparison of these spaces is interesting, a main purpose of the global view is to enable the selection of instances, from which to explore the local explanations. The projection attribution of the Primary Instance (PI) is explored and typically viewed with an optional Comparison Instance (CI).

@@

The global view facilitated the selection instances. The feature-level attribution of the primary instance is normalized and used as the initial 1D basis in a radial tour. This is an approximation of the contributions of the linear features that best explain the difference between the model intercept and an instance's prediction, not the local shape of the model surface.

The initial frame is the normalized SHAP values of the PI. The current projection basis is depicted as the width of a bar, the feature's contribution to the horizontal axis. The normalized values of all instances are shown as vertical parallel coordinate plots.

The radial tour creates a basis path by varying the contribution of a selected feature, fully into and out of a projection frame. Doing so tests an individual feature's sensitivity to the structure identified by the local explanation. The default feature selected has the largest discrepancy between the attribution of primary and comparison instances. The following sections elaborate on the takeaways from applying this approach in classification and regression tasks. Now that we have introduced the global view and corresponding cheem radial tour, let us discuss the differences between the classification and regression cases.

@@

Typically we select a misclassified instance compared to a correctly classified point nearby in data space. The model infomation in the global view is a depiction of the model conmfusion matrix. The radial tour is 1D, with density display, while the goal of the tour; exploring the sensitivity of each feature to structure identified by the local explanation, evaluating the support or robustness of the prediction.

@@

\caption{Overview of the cheem viewer for classification tasks. Plots are colored on predicted class, and red circles indicate misclassified instances. Model visual is essentially a representation of the confusion matrix of the model. The radial tour is a 1D projection starting the attribution projection. The first frame is the linear feature importances that best describe the instance's prediction. We probe the support of feature contributions by selecting a feature and use the radial tour to vary its coefficient.}\label{fig:classificationcase}
\end{figure}

@@

A tooltip displays instance number/name and classification information while the cursor hovers over a point. Linked brushing allows the selection of points (left click and drag) where those points will be highlighted across plots. The information corresponding to the selected points is populated on a dynamic table. These interactions aid exploration of the spaces and, finally, identification of a primary and comparison instance.

@@

\caption{Illustration of data explorations interactions in the global view. This view has linked brushing, where instances selected in one facet are highlighted in the other facets and populate an interactive tabular display below. Tooltips display when hovering over an instance.}\label{fig:interactions}
\end{figure}

@@

\begin{itemize}
    \item \textbf{Data:} predictors and response are unscaled complete numerical matrix. Most models and local explanations are scale invariant.
    \item \textbf{Model and explanation:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [Liaw and Wiener (2002)], compatibility tree SHAP. We use modest hyperparameters, namely: 125 trees, number features randomly sampled at each split, mtry = $\sqrt{p}$ or $p/3$ for classification and regression, and minimum size of terminal nodes $max(1, n/500)$ or $max(5, n/500)$ for classification and regression. Tree SHAP is calculated for \emph{each} instance using the package \textbf{treeshap} Kominsarczyk et al. (2021). This implementation aggregates over exhaustively over the attribution of all trees and we opt to not to fit interactions of features. 
    \item \textbf{Cheem view:} after the model and full explanation space are calculated we scale each feature by standard deviations away from the mean to achieve a common support for visuals. Statistics for mapping to color are calculated on the scaled spaces. Interactive tabular display reports the original values.
\end{itemize}

The time to preprocess the data will vary significantly with the model and local explanation. For reference, the FIFA data, 5000 instances of nine explanatory features, took 2.5 seconds to fit a random forest model of modest hyperparameters. Extracting the tree SHAP values of each instance took 270 seconds combined. PCA and statistics of the features and attributions took 2.8 seconds. These runtimes were from a non-parallelized R session on a modern laptop, but suffice to say that the bulk of the time will be spent on the local attribution. An increase in model complexity or data dimensionality will quickly become an obstacle. With its reduced computational complexity, this makes tree SHAP a good candidate to start with. Alternatively, the package \textbf{fastshap} (Greenwell 2020) claims extremely low runtimes, which are attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting.

@@

{\centering \includegraphics[width=1\linewidth]{./figures/case_penguins_BlFl}

\caption{An orthogonal view of penguins data showing the PI, a Gentoo (purple) close to the distribution of Chinstrap (orange).}\label{fig:casepenguinsblfl}
\end{figure}

The Palmer penguins data (Gorman, Williams, and Fraser 2014; Horst, Hill, and Gorman 2020) was collected on three species of penguins foraging near Palmer Station, Antarctica. The data was publicly available to be a substitute for the overly-used iris data, and is quite similar in form. After removing incomplete observations, there are 333 instances and we will use the four physical measurements, \texttt{bill\_length\_mm} (\texttt{b\_l}), \texttt{bill\_depth\_mm} (\texttt{b\_d}), \texttt{flipper\_length\_mm} (\texttt{f\_l}), \texttt{body\_mass\_g} (\texttt{wgt}), for this illustration. A random forest model was fitted with species as the response feature.

@@

\caption{Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is an Chinstrap (orange) penguin that is misclassified as a Gentoo (purple), marked as an asterisk in (a), and the dashed vertical line in (b). Starting from the attribution projection the contribution of \texttt{b\_l} is varied. This explanation is not supported sense by the time there is moderate contribution. The basis showing the correct species separation is fairly close to the regularly attribution to other Gentoo penguins aside from the sign change on flipper length. The .mp4 animation of this tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{shorturl.at/hFHNS}}\label{fig:casepenguins}
\end{figure}

Figure \ref{fig:casepenguins} shows plots from the cheem viewer for exploring the random forest model on the penguins data. Plot (a) shows the global view, and plot (b) shows several 1D projections generated with the radial tour. Penguin 243, an Chinstrap (orange), is the PI because it has been misclassified as a Gentoo (purple).

There is more separation visible in the attribution space than the data space, as would be expected. The observed vs predicted plot reveals a handful of misclassified instances. A misclassified point (represented by the asterisk in the global view, and as a dashed vertical line in the tour view) is contrasted with a correctly classified point (represented by an \(\times\) and vertical dotted line).

The radial tour starts from the attribution projection of the misclassified instance (left plot). The important features identifed by SHAP in the (wrong) prediction for this instance is mostly \texttt{b\_l} and \texttt{b\_d} with small contributions of \texttt{f\_l} and \texttt{wgt}. This projection is a view where the Gentoo (purple) looks much more likely for this instance than Chinstrap. For this instance we would like to know why an Adelie has been wrongly labelled a Chinstrap, so this combination of features is not particularly useful. To explore this we use the radial tour to vary the contribution of flipper length (\texttt{f\_l}). The more that \texttt{f\_l} contributes to the the projection the more this penguin looks like a Gentoo, its observed species. This suggests that \texttt{f\_l} should have been an important feature for explaining the (wrong) prediction.

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales across nine features. Using interaction from the global view, we select a house with an extreme negative residual and an accurate instance close to it in the data.
